\documentclass[twoside,english]{uiofysmaster}

\usepackage{preamble}
\usepackage{subfiles}
\usepackage{datetime}
\newdateformat{monthyeardate}{%
  \monthname[\THEMONTH], \THEYEAR}


\author{Bendik Samseth}
\title{(Title Tinkering Needed)Bootstrapping Variational Monte Carlo with Machine Learning}
\date{\monthyeardate\today}

\input{glossary.tex}
\makeglossaries

\begin{document}

\maketitle

\begin{abstract}
Recent applications of machine learning for quantum mechanics have shown
encouraging results in efforts to overcome the exponential scaling complexity of
the many-body wave function. We continue this exploration by introducing a
neural network as an additional Jastrow factor to the wave function ansatz
within the framework of \acrlong{vmc}, with the hope of learning
correlations beyond what traditional methods have been able to.

We begin with a review of the relevant parts of quantum mechanics with particular
emphasis on \acrlong{mc} methods. We then introduce central elements of machine
learning, focusing on neural networks. Finally we discuss implementation
challenges.

We test our approach first on two interacting electrons in a harmonic oscillator
potential. Where the benchmark estimates $\expval{E_0}=\SI{3.00064(4)}{\au}$, we
outperform this by a lowering of more than an order of magnitude, resulting in
$\expval{E_0}=\SI{3.000021(4)}{\au}$. The energy estimates approach those of
\acrlong{dmc}, which can produce exact solutions to the Schrödinger equation.

Second we apply the same technique to the strongly correlated system of liquid
$^4$He. Again we obtain significant lowering of the ground state energy, from
the benchmark result of $\expval{E_0}=\SI{6.76(2)}{\kelvin}$ to
$\expval{E_0}=\SI{6.96(2)}{\kelvin}$, although the optimization problem proves far
more challenging.

These improvements come at the cost of greatly increased computing time,
however, we argue that the time complexity can still be made to scale similarly to
$\mathcal{O}(N^2)$ with the number of particles.
\end{abstract}

\glsresetall   % Accronyms will be defined in main text, not in abstract.

\begin{acknowledgements}
  I acknowledge my acknowledgements.
\end{acknowledgements}

\tableofcontents

\listoffigures
\begingroup
\let\clearpage\relax
\listoftables
\endgroup

\subfile{Introduction.tex}

\part{Theory}
\label{prt:theory}
\subfile{QuantumTheory.tex}
\subfile{VariationalMonteCarlo.tex}
\subfile{MonteCarlo.tex}
\subfile{MachineLearning.tex}
\subfile{MergingVMCandML.tex}
\part{Implementation}
\label{prt:implementation}
\subfile{Parallelization.tex}
\subfile{AutomaticDifferentiation.tex}
\subfile{Qflow.tex}
\subfile{Verification.tex}
\part{Results}
\label{prt:results}
\subfile{QuantumDots.tex}
\subfile{LiquidHelium.tex}
\part{Conclusion and Outlook}
\label{prt:conclusion}
\subfile{Conclusion.tex}
\subfile{Appendix.tex}
\printglossary[title=List of Accronyms, toctitle=Terms and abbreviations,type=\acronymtype]
% \begingroup
% \let\clearpage\relax
% \printglossary[title=List of Terms]
% \endgroup

\printbibliography[heading=bibintoc, title={References}]


\end{document}
